An active sampling model of mental rotation

### 2014-01-19 man/cogsci-2014/cogsci-2014.tex
More work on modeling section

\newcommand{\TODO}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\Xa}[0]{\mathbf{X}_a}
\newcommand{\Xb}[0]{\mathbf{X}_b}
\newcommand{\Xt}[0]{\mathbf{X}_t}
\newcommand{\R}[0]{\mathbf{R}_\theta}
\newcommand{\F}[0]{\mathbf{F}}
\newcommand{\M}[0]{\mathbf{M}}
\newcommand{\I}[0]{\mathbb{I}}
\newcommand{\hi}[0]{h=0}
\newcommand{\hf}[0]{h=1}

@@

Formally, we denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is
generated by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$,
where $\theta$ is a rotation, $h=0$ is the hypothesis that the images
depict the same object, and $h=1$ is the hypothesis that the images
depict mirror-image objects. The posterior probability of each
hypothesis given the observed shapes is then:
\begin{equation}
  p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta, h)p(X_a)p(h)p(\theta)\ \mathrm{d}\theta.
  \label{eq:posterior}
\end{equation}
Because we ultimately want to determine which hypothesis is more
likely, the quantity of interest is a ratio:
\begin{equation*}
  \ell := \frac{\int p(X_b\ \vert\ X_a, \theta, \hi)p(X_a)p(\hi)p(\theta)\ \mathrm{d}\theta}{\int p(X_b\ \vert\ X_a, \theta, \hf)p(X_a)p(\hf)p(\theta)\ \mathrm{d}\theta},
\end{equation*}
which (assuming the two hypotheses are equally likely \textit{a
  priori}, and that all rotations are equally likely) simplifies to:
\begin{equation}
  \ell = \frac{\int p(X_b\ \vert\ X_a, \theta, \hi)\ \mathrm{d}\theta}{\int p(X_b\ \vert\ X_a, \theta, \hf)\ \mathrm{d}\theta}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell > 1$, then we accept the hypothesis that the images depict
the same object ($\hi$); if $\ell < 1$, then we accept the hypothesis
that the images depict flipped objects ($\hf$).

@@

We define $p(\Xb\ \vert\ \Xa, \theta, h)$ to be the similarity between
$\Xb$ and a transformation of $\Xa$:
\begin{equation}
  p(\Xb\ \vert\ \Xa, \theta, h):= S(\Xb, f(\Xa, h, \theta)).
  \label{eq:likelihood}
\end{equation}
We do not know which vertices of $\Xb$ correspond to which vertices of
$\Xa$, so $S$ must marginalize over the set of possible mappings,
$\mathbb{M}$. For brevity, let $\mathbf{X}_m=\M\cdot{}f(\Xa, h,
\theta)$ where $\M\in\mathbb{M}$ is a permutation matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{|\mathbb{M}|} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma^2)
  \label{eq:similarity}
\end{equation}

Additionally, we have the constraint of small
transformations. Specifically, if the current mental image is $\Xt$,
then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, \epsilon) &\mbox{ rotate right by $\epsilon$,} \\
      f(\Xt, 0, -\epsilon) &\mbox{ rotate left by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip.} \\
    \end{array} \right.
\end{equation}

\subsubsection{Hill climbing model}

An intuitive and simple model is to first rotate in the direction that
increases similarity, and then stop rotating once a ``match'' is
found. If all rotations have been exhausted, then flip, and try
rotating again.

\subsubsection{Bayesian quadrature model}

>>>>

\newcommand{\TODO}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\Xa}[0]{\mathbf{X}_a}
\newcommand{\Xb}[0]{\mathbf{X}_b}
\newcommand{\Xt}[0]{\mathbf{X}_t}
\newcommand{\R}[0]{\mathbf{R}_\theta}
\newcommand{\F}[0]{\mathbf{F}}
\newcommand{\M}[0]{\mathbf{M}}
\newcommand{\I}[0]{\mathbb{I}}
\newcommand{\hi}[0]{h=0}
\newcommand{\hf}[0]{h=1}
\newcommand{\dif}[0]{\,\mathrm{d}}

@@

Formally, we denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is
generated by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$,
where $\theta$ is a rotation, $h=0$ is the hypothesis that the images
depict the same object, and $h=1$ is the hypothesis that the images
depict mirror-image objects. The posterior probability of each
hypothesis given the observed shapes is then:
\begin{equation}
  p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta, h)p(X_a)p(h)p(\theta)\dif\theta.
  \label{eq:posterior}
\end{equation}
Because we ultimately want to determine which hypothesis is more
likely, the quantity of interest is a ratio:
\begin{equation*}
  \ell := \frac{\int p(X_b\ \vert\ X_a, \theta, \hi)p(X_a)p(\hi)p(\theta)\dif\theta}{\int p(X_b\ \vert\ X_a, \theta, \hf)p(X_a)p(\hf)p(\theta)\dif\theta},
\end{equation*}
which (assuming the two hypotheses are equally likely \textit{a
  priori}, and that all rotations are equally likely) simplifies to:
\begin{equation}
  \ell = \frac{\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta}{\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell > 1$, then we accept the hypothesis that the images depict
the same object ($\hi$); if $\ell < 1$, then we accept the hypothesis
that the images depict flipped objects ($\hf$).

@@

We define $p(\Xb\ \vert\ \Xa, \theta, h)$ to be the similarity between
$\Xb$ and a transformation of $\Xa$:
\begin{equation}
  p(\Xb\ \vert\ \Xa, \theta, h):= S(\Xb, f(\Xa, h, \theta)).
  \label{eq:likelihood}
\end{equation}
We do not know which vertices of $\Xb$ correspond to which vertices of
$\Xa$, so $S$ must marginalize over the set of possible mappings,
$\mathbb{M}$. For brevity, let $\mathbf{X}_m=\M\cdot{}f(\Xa, h,
\theta)$ where $\M\in\mathbb{M}$ is a permutation matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{|\mathbb{M}|} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma^2).
  \label{eq:similarity}
\end{equation}

Additionally, we have the constraint of small
transformations. Specifically, if the current mental image is $\Xt$,
then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, \epsilon) &\mbox{ rotate right by $\epsilon$,} \\
      f(\Xt, 0, -\epsilon) &\mbox{ rotate left by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip.} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}

\subsubsection{Threshold-based model}

An intuitive and simple model is to first rotate in the direction that
increases similarity, and then stop rotating once a ``match'' is
found, as determined by a threshold on the value of $S$. If all
rotations have been exhausted, then flip, and try rotating again.

The consequence of doing a search for a match is that
Eq.~\ref{eq:lh-ratio} becomes a likelihood ratio test, where $\theta$
is set to the MLE value, rather than being marginalized:
\begin{equation}
  \ell = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)}.
  \label{eq:mle-lh-ratio}
\end{equation}

\subsubsection{Hill-climbing model}

Unfortunately, it is not always clear what constitutes a good enough
``match''. Many shapes may have multimodal similarity functions
\TODO{refer to figure}, and the maximum possible value of $S$ is
unlikely to be known \textit{a priori} \TODO{cite something about how
  shape matching is hard}. In this case, we must weaken our stopping
criteria: rather than searching for a value above a threshold, we
search for \textit{any} local maximum.  Similar to the threshold-based
model, we use the (local) MLE estimate of $\theta$, as in
Eq.~\ref{eq:mle-lh-ratio}.

\subsubsection{Confidence-based model}

An alternative to the previous two models is to approximate the
integrals in Eq.~\ref{eq:lh-ratio} by picking locations to sample
which will most improve accuracy of the estimate. Rather than a point
estimate of the ratio, or even a simple Monte Carlo estimate, we can
infer a distribution by placing priors over the similarity functions
themselves:
\begin{equation}
\ell = \frac{\int \left[\int S(\Xb, f(\Xa, \theta, \hi))p(\theta)\dif\theta\right] p(S)\dif S}{\int \left[\int S(\Xb, f(\Xa, \theta, \hf))p(\theta)\dif\theta\right] p(S)\dif S}
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use Gaussian Process (GP) priors \cite{Osborne:2012tm} \TODO{probably
  need to describe better what a GP is}. Because our domain is on a
circle, we utilize a periodic kernel for the covariance
\cite{Rasmussen:2006vz}, i.e. $k(\theta,
\theta^\prime)=h^2\exp[-\frac{2}{w^2}\sin^2(\frac{1}{2}(\theta-\theta^\prime))]$.

Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we first place a GP prior on
$\log(S_h)$. Because GPs have no guarantee of non-negativity, placing
the prior over the log enforces positivity after it is exponentiated:
\begin{equation}
\mathbb{E}[Z_h] \approx \int \exp(\mu_h(\theta))p(\theta)\dif\theta,
\end{equation}
where $\mu_h:=\mu(\log S_h)$ is the mean of the log-GP. However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP.\footnote{Note that the approach
  presented in \cite{Osborne:2012tm} differs slightly from the one
  used here. We chose this newer version after a discussion with one
  of the authors on the original BQ paper \TODO{cite personal
    communication}.}We will denote these points as
$\bar{S}_h:=\exp(\mu_h)$. Then, from \TODO{cite personal
  communication}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

Now that we have a distribution over $Z_h$, we can additionally
compute which points which we would expect to strengthen our estimate
(i.e., decrease the variance). From \cite{Osborne:2012tm}, the
expected variance of $Z_h$ if we sampled a new observation at
$\theta_a$ is:
\begin{align}
\mathbb{E}&[\mathbb{V}(Z_h|\theta_a)]=\mathbb{V}(Z_h) + \mathbb{E}[Z_h] - \\
&\int \mathbb{E}[Z_h|\theta_{a}]^2 \mathcal{N}(\mu_h(\theta_a), \mathrm{Cov}_h(\theta_a, \theta_a))\dif\log S_h(\theta_a)\nonumber
\label{ref:expected-variance}
\end{align}

Revisiting our likelihood ratio, we now have:
\begin{equation}
p(\ell)\approx\frac{\mathcal{N}(\mathbb{E}[Z_0], \mathbb{V}(Z_0))}{\mathcal{N}(\mathbb{E}[Z_1], \mathbb{V}(Z_1))}
\end{equation}
\TODO{figure out what to do here.}
